Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-11-24 survey/survey.tex
add some text to a section, and then comment out the whole section as it is not relevant anymore. Checking this in to preserve it, but will actually delete it next.

\documentclass{article}
\usepackage{fullpage}
\usepackage{url}
\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan P. James, Amr Sabry}

@@

Many resolutions were offered for this conundrum (for a compilation, see
Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
principle whose experimental validation
happened recently in 2012~\cite{lutz2012}.

@@

And that brings us to the objective of this survey: explore the domain of
\emph{typed reversible computation}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Models of Computation}

Explain classical models of computation and their limitations in
real-world ``physical'' applications.

Computation in the ``real world'' vs. idealized computation in
abstract models.
\begin{itemize}
\item Mathematical abstraction of sorting is well-understood; sorting
  petabytes of data violates many of the assumptions: hardware
  component failure frequent, global synchronization impossible;
  consistency must be eventual; etc.
\item Mathematical abstraction of sortedness property: easy; in the
  real world with petabytes of evolving data; it is not clear that it
  is even possible to check sortedness; easy to get inconsistent
  results
\item Noise does not exist in our mathematical abstractions of
  computation; essential use in differential privacy
\item Mathematical models of secure computing typically ignore power
  consumption, electromagnetic signatures; actual attacks based on DPA
  and DEMA however
\end{itemize}

% \jc{While all the above are true (and not exhaustive), what I also read
% is that the first two are actually ``the same'' point mostly related
% to scale and distributivity, the 3rd is about inadequate models (but not
% related to physics), while the 4th is about the physics of devices.
% They are not about CPS. And not about Quantum either. So how are these
% motivating a deeper look at how quantum affects our MoC? It feels to me
% that we ought to have a quantum ``example'' here already.}

% \jc{What this seems to motivate is to look at our models with respect
% to what we're trying to achieve.  Models are, by definition, abstractions
% that ``abstract away'' irrelevant details. So the point would be that
% some details that we thought were irrelevant really are not.
% So I could see this section motivating a whole programme of
% targeted de-abstraction. This could lead into the focus of this paper
% following one strand, namely one aspect of quantum physics.}

\begin{small}
\begin{quote}
``Turing hoped that his abstracted-paper-tape model was so simple, so
  transparent and well defined, that it would not depend on any
  assumptions about physics that could conceivably be falsified, and
  therefore that it could become the basis of an abstract theory of
  computation that was independent of the underlying physics.  `He
  thought,' as Feynman once put it, `that he understood paper.' But he
  was mistaken.''
\end{quote}
\hfill --- \emph{The Fabric of Reality}~\cite{deutsch}
\end{small}

@@

We begin with a bold thesis: \textit{common computational models
  embody various physical assumptions that are plainly
  \emph{wrong}}. As we discuss in the remainder of this chapter, we
are not the first to make such a statement. We do, however, make two
novel contributions:
\begin{itemize}
\item we argue on pragmatic terms that our physically incorrect
  computational abstractions contribute significantly to the
  ``software crisis'' and that they inhibit advances in new emerging
  application domains; 
\item we identify the physical principle of \textbf{conservation of
  information} as a simple twist to our computational models that has,
  as we demonstrate in this book, far reaching theoretical and
  practical positive consequences to the science of computing.
\end{itemize}

To appreciate the first point, we examine various textbook abstractions and
contrast them with the actual (i.e., physical) computational situations to
which they correspond. Generally speaking, we find that textbook algorithms
differ from ``real programs'' in their assumptions about their environment of
operation. In textbook abstractions, the surrounding environment which
includes, for example, the data on which the algorithm operates is considered
static and well-defined. In the real world, this environment is dynamic and
noisy.

\paragraph*{Sorting.} 
One might consider sorting to be a fairly simple ``solved'' problem. Indeed
textbook sorting algorithms have been formally specified, proved correct,
analyzed in the best case, worst case, average case, and are even taught in
many high school curricula. A beginner programmer can readily implement such
algorithms to sort ``real data'' --- or can he?

Under some assumptions, the answer is clearly yes. However, a moment's
reflection should convince us that the textbook algorithm is too idealized
and that sorting real data in the real world is still a challenge!  Consider
the following quote posted on the Google research blog in 2011:
\begin{quote}
  Almost three years ago we announced results of the first ever ``petasort''
  \ldots . It completed in just over six hours on 4000 computers. Recently we
  repeated the experiment using 8000 computers. The execution time was 33
  minutes, an order of magnitude improvement~\cite{petasort}.
\end{quote}
What happens at the petabyte scale that makes sorting such a
challenge? The answer is simply that ``the real world happens!'' In
the real world, many of the idealized mathematical assumptions
embodied in our computational models are invalidated. For example,
proper hardware functioning can no longer be taken for granted as
component failure becomes the norm, global synchronization becomes
impossible, determinism becomes expensive to maintain, speculative
execution becomes necessary, global consistency becomes impractical to
maintain, and eventual consistency is accepted; etc. There have been
many attempts to taken some of these issues into account in various
computational models but these attempts are arguably not going far
enough by not questioning the physical assumptions in our
computational models and re-thinking our very foundations.

\paragraph*{Binary Search.} 
In the idealized textbook setting, the binary search algorithm has the
precondition that the data is sorted. Even if we just have a few thousands
items, na\"\i vely checking sortedness (in linear time) defeats the purpose of
the efficient (logarithmic) binary search. If we have petabytes of
(changing!) data, is it even possible, in principle, to check sortedness? In
fact, in this dynamic situation, one is likely to get inconsistent
results. In other words, the very idea of pre-conditions and post-conditions
that can be checked statically appears to be at odds with computing in the
``real world.''

@@

\paragraph*{A Crisis?}
Is this tension between theory and practice a crisis?  We argue:
yes. Computer science theory is founded on abstractions of the real
world. These abstractions are necessary to shield us from details of
technology, noise, errors, etc. and they have served us well until
recently.  It is, however, becoming more and more evident that
computer science theory is becoming less and less relevant to
practice, and hence it is important to revisit our abstractions from a
foundational point of view.

%%%
\subsection{Our Origins}

@@

One of these wild claims must be true!:
\begin{itemize}
\item the extended Church-Turing thesis is false, or
\item quantum physics is false, or
\item there is an efficient classical algorithm for factoring
\end{itemize} 
Indeed, if quantum physics is correct then there is an efficient quantum
algorithm for factoring (Shor). If there is no efficient classical algorithm
for factoring then the extended Church-Turing thesis is false.

Indeed, conventional classical models of computation, including boolean
logic, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

The irreversible abstractions we choose as the basis of our most
primitive models of computing are at ``odds'' with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through the remarkable \emph{Landauer
  principle}~\cite{Landauer:1961}. Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
  physical quantity} in much the sense that one would think of an
electron as a physical quantity, and that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
as mass and energy are.

@@

%subcode{code}
%! language = Haskell
% check-password (guess) = 
%   if guess == "10"
%   then True
%   else False

@@

%%%
\subsection{Equalities as Isomorphisms}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}

Perhaps Information Flow Security

% \jc{This is where I am still least sure. I am wary of either doing
% new research with such a tight timeline, or of being too speculative.
% If there are papers on this already, perhaps references to them could
% be put here?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}
\pagestyle{headings}
\bibliography{cites}
\end{document}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}

\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan P. James, Amr Sabry}

@@

Many resolutions were offered for this conundrum (for a compilation, see
Maxwell's Demon~\cite{leff1990}), but none withstood careful
scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
principle whose experimental validation
happened recently in 2012~\cite{berut2012experimental}.

@@

And that brings us to the principal objective of this survey: explore the
domain of \emph{typed reversible computation}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% JC: I am commenting this out as, intereting as this all is, a lot of
%     it is about ``real world'' and NOT about the intersection of physics
%     and computation.  It would be possible to write about that too.
%     For example, the hardware failures of petabyte scale computing is
%     physical; physics-based attacks on privacy and security; physics
%     of dealing with noisy environments (key debouncing, timing drifts,
%     spurious data, heart rate spikes, etc);
\begin{comment}
\section{Physics and computation}

@@

We begin with a bold thesis: \textit{common computational models
embody various physical assumptions that are plainly
\emph{wrong}}. We certainly
are not the first to make such a statement. We do want to point
out that these physically incorrect computational abstractions
inhibit advances in computing in general.

To appreciate this point, we examine various textbook abstractions and
contrast them with the actual (i.e., physical) computational situations to
which they correspond. Generally speaking, we find that textbook algorithms
differ from ``real programs'' in their assumptions about their environment of
operation. In textbook abstractions, the surrounding environment which
includes, for example, the data on which the algorithm operates is considered
static and well-defined. In the real world, this environment is dynamic and
noisy.

\paragraph*{Sorting.} 
One might consider sorting to be a fairly simple ``solved'' problem. Indeed
textbook sorting algorithms have been formally specified, proved correct,
analyzed in the best case, worst case, average case, and are even taught in
many high school curricula. A beginner programmer can readily implement such
algorithms to sort ``real data'' --- or can she?

Under some assumptions, the answer is clearly yes. However, a moment's
reflection should convince us that the textbook algorithm is too idealized
and that sorting real data in the real world is still a challenge!  Consider
the following quote posted on the Google research blog in 2011:
\begin{quote}
  Almost three years ago we announced results of the first ever ``petasort''
  \ldots . It completed in just over six hours on 4000 computers. Recently we
  repeated the experiment using 8000 computers. The execution time was 33
  minutes, an order of magnitude improvement~\cite{petasort}.
\end{quote}
What happens at the petabyte scale that makes sorting such a
challenge? The answer is simply that ``the real world happens!'' In
the real world, many of the idealized mathematical assumptions
embodied in our computational models are invalidated. For example,
proper hardware functioning can no longer be taken for granted, as
component failure becomes the norm, global synchronization becomes
impossible, determinism becomes expensive to maintain, speculative
execution becomes necessary, global consistency becomes impractical to
maintain, and eventual consistency is accepted; etc. There have been
many attempts to taken some of these issues into account in various
computational models but these attempts are arguably not going far
enough by not questioning the physical assumptions in our
computational models and re-thinking our very foundations.

As another data point to reinforce this point: What algorithm
does Python (and other mainstream platforms) use? A variant of
QuickSort or HeapSort perhaps.  No. Timsort. This hybrid algorithm
performs rather well in practice -- but is complex, and even harder 
to formally analyze. It does not tend to get mentioned in algorithms
textbooks at all.

\paragraph*{Binary Search.} 
In the idealized textbook setting, the binary search algorithm has the
precondition that the data is sorted. Even if we just have a few thousands
items, na\"{\i}vely checking sortedness (in linear time) defeats the purpose of
the efficient (logarithmic) binary search. If we have petabytes of
(changing!) data, is it even possible, in principle, to check sortedness? 
Or guarantee sortedness? In
fact, in this dynamic situation, one is likely to get inconsistent
results. In other words, the very idea of pre-conditions and post-conditions
that can be checked statically appears to be at odds with computing in the
``real world.''

@@

\paragraph*{A Crisis?}
Is this tension between theory and practice a crisis?  We argue:
yes. Computer science theory is founded on abstractions of the real
world. These abstractions are necessary to shield us from details of
technology, noise, errors, etc. and they have served us well until
recently.  It is, however, becoming more and more evident that
computer science theory is becoming less and less relevant to
practice, and hence it is important to revisit our abstractions from a
foundational point of view.
\end{comment}

\section{Models of Computation}
%%%
\begin{comment}
\subsection{Our Origins}

@@

One of these wild claims must be true!:
\begin{itemize}
\item the extended Church-Turing thesis is false, or
\item quantum physics is false, or
\item there is an efficient classical algorithm for factoring
\end{itemize} 
Indeed, if quantum physics is correct then there is an efficient quantum
algorithm for factoring (Shor). If there is no efficient classical algorithm
for factoring then the extended Church-Turing thesis is false. 
\end{comment}

Conventional classical models of computation, including boolean
logic, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

These irreversible abstractions chosen as the basis of our
models of computing are at ``odds'' with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through the remarkable \emph{Landauer
  principle}~\cite{Landauer:1961}. Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity, and that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
as mass and energy are.

@@

% verbatim for now, will go to something decent later
\begin{verbatim}
check-password (guess) = 
  if guess == "10"
  then True
  else False
\end{verbatim}

@@

%%%
\section{Reversibility}

\subsection{Equalities as Isomorphisms}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}
%\pagestyle{headings}
\bibliography{cites}
\end{document}
